\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-05-19 paper.tex
update beta dependence

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
\label{table:shared_associations}
\end{table}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\beta_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:beta-update}
  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{equation}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{equation}
for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.

Concerning the update of ${\gamma_k^0}^{(w)}$ and $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k^0, \gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k^0$ and $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}. The update for ${\gamma_k^0}^{(w)}$ therefore requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k^0}
  {\gamma_k^0}^{(w+1)} \in \argmin_{\gamma_k^0 \in \R^p} Q^{(w)}_{n,k}(\gamma_k^0, {\gamma_k}^{(w)}) + \zeta_{1,k} \norm{\gamma_k^0}_{\text{en}, \eta}.
\end{equation}
We then rewrite problem~\eqref{eq:minimization-gamma_k^0} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-gamma_k^0}
  \begin{split}
    \text{minimize}& \quad \quad Q_{n, k}^{(w)}({\gamma_k^0}^+ - {\gamma_k^0}^-, {\gamma_k}^{(w)}) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p ({\gamma_{k,j}^0}^+ + {\gamma_{k,j}^0}^-) + \dfrac \eta 2 \norm{{\gamma_k^0}^+ - {\gamma_k^0}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad {\gamma_{k,j}^0}^+ \geq 0 \text{ and } {\gamma_{k,j}^0}^- \geq 0 \text{ for } j = 1, \ldots, p.
  \end{split} 
\end{equation}
The gradient is here given by
\begin{align}
  \label{eq:grad-Q-1}
  \frac{\partial Q_{n,k}^{(w)}(\gamma_k^0, {\gamma_k}^{(w)})}{\partial \gamma_k^0} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \nonumber \\
  &\times \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, {\gamma_k}^{(w)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i.
\end{align}
Then, the update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
We treat problem~\eqref{eq:minimization-gamma_k} in the same way as~\eqref{eq:minimization-beta_k}, that is
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \cdot), \tilde \eta, \zeta_{3,k}\big)
\end{equation}
with the gradient being known since
\begin{align}
  \label{eq:grad-Q-2}
  &\frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w+1)}, \gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w+1)}, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

@@

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
  Update $({\gamma_k^0}^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k^0} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}
Let us define the relative difference at step $(w+1)$ as
\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
Then in practice, the convergence for Algorithm~\ref{alg:QNMCEM} is declared when the condition
\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2 \]
is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.

@@

In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, \hat \beta^l, b_i^l) \Big\}.\]

>>>>

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_a(t, b^l)$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
\label{table:shared_associations}
\end{table}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\beta_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:beta-update}
  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{equation}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{equation}
for all $l = 1, \ldots, L$.

Concerning the update of ${\gamma_k^0}^{(w)}$ and $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k^0, \gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k^0$ and $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k^0, \gamma_k}^1$ defined in~\eqref{eq:g-tilde-1-2}. The update for ${\gamma_k^0}^{(w)}$ therefore requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k^0}
  {\gamma_k^0}^{(w+1)} \in \argmin_{\gamma_k^0 \in \R^p} Q^{(w)}_{n,k}(\gamma_k^0, {\gamma_k}^{(w)}) + \zeta_{1,k} \norm{\gamma_k^0}_{\text{en}, \eta}.
\end{equation}
We then rewrite problem~\eqref{eq:minimization-gamma_k^0} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-gamma_k^0}
  \begin{split}
    \text{minimize}& \quad \quad Q_{n, k}^{(w)}({\gamma_k^0}^+ - {\gamma_k^0}^-, {\gamma_k}^{(w)}) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p ({\gamma_{k,j}^0}^+ + {\gamma_{k,j}^0}^-) + \dfrac \eta 2 \norm{{\gamma_k^0}^+ - {\gamma_k^0}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad {\gamma_{k,j}^0}^+ \geq 0 \text{ and } {\gamma_{k,j}^0}^- \geq 0 \text{ for } j = 1, \ldots, p.
  \end{split} 
\end{equation}
The gradient is here given by
\begin{align}
  \label{eq:grad-Q-1}
  \frac{\partial Q_{n,k}^{(w)}(\gamma_k^0, {\gamma_k}^{(w)})}{\partial \gamma_k^0} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \nonumber \\
  &\times \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, {\gamma_k}^{(w)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i.
\end{align}
Then, the update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
We treat problem~\eqref{eq:minimization-gamma_k} in the same way as~\eqref{eq:minimization-beta_k}, that is
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \cdot), \tilde \eta, \zeta_{3,k}\big)
\end{equation}
with the gradient being known since
\begin{align}
  \label{eq:grad-Q-2}
  &\frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w+1)}, \gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w+1)}, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

@@

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
  Update $({\gamma_k^0}^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k^0} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}
Let us define the relative difference at step $(w+1)$ as
\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
Then in practice, the convergence for Algorithm~\ref{alg:QNMCEM} is declared when the condition
\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.

@@

In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, b_i^l) \Big\}.\]
